CrypTool Script \\ Mathematics and Cryptography

### 2003-03-11 script/en/chapters/primes.tex
be updates

Prime numbers are particularly important for 3 reasons:
\begin{itemize}
  \item In number theory, they are considered to be the basic components of
natural numbers, upon which numerous brilliant mathematical ideas are based.
  \item They are of extreme practical importance in modern
\index{Cryptography!modern} cryptography (public key \index{Cryptography!public key} cryptography). The most common public key procedure, invented at the end of
the 1970's, is \index{RSA} RSA encryption. Only using (large) prime numbers for
particular parameters can you guarantee that an algorithm is secure, both for
the RSA procedure and for even more modern procedures (digital
\index{Signatures!digital} signature, elliptic curves).
  \item The search for the largest known prime numbers does not have any
practical usage known to date, but requires the best computers, is an excellent
benchmark (possibility for determining the performance\index{Performance} of computers) and leads
to new calculation methods on many computers \\ (see also:
\href{http://www.mersenne.org/prime.htm}{\tt
http://www.mersenne.org/prime.htm}).
\end{itemize}
Many people have been fascinated by prime numbers over the past two millenniums.
Ambition to make new discoveries about prime numbers has often resulted in
brilliant ideas and conclusions. The following section provides an easily
comprehensible introduction to the basics of prime numbers. We will also explain
what is known about the distribution (density, number of prime numbers in
particular intervals) of prime numbers and how prime number tests work.

@@

The sequence of prime numbers starts with $$ 2,~ 3,~ 5,~ 7, ~ 11, ~ 13, ~
17, ~ 19, ~ 23, ~ 29, ~ 31, ~ 37, ~ 41, ~ 43, ~ 47, ~ 53, ~ 59, ~ 61,
~ 67, ~ 71, ~ 73, ~ 79, ~ 83, ~ 89, ~ 97, \cdots . $$
The first 100 numbers include precisely 25 prime numbers. After this, the percentage of primes
constantly decreases. Prime numbers can be factorised in a unique {\em trivial}
way: $$5 = 1 \cdot 5,\quad  17 = 1 \cdot 17, \quad 1,013 = 1 \cdot 1,013,  \quad
1,296,409 = 1 \cdot 1,296,409.$$
All numbers that have $2$ or more factors are called \index{Number!composite}
{\em composite}. These include $$ 4 = 2 \cdot 2, \quad 6 = 2\cdot 3 $$ as well
as numbers that {\em look like primes}, but are in fact composite:
$$ 91 = 7 \cdot 13, \quad 161=7 \cdot 23, \quad 767 =13 \cdot 59. $$

@@

Although there are only $92$ natural chemical elements, the number of prime
numbers is unlimited. Even the Greek, \index{Euclid} Euclid%
\footnote{Euclid,
a Greek mathematician of 4th and 3rd century before Christ. He worked at the
Egyptian academy of Alexandria and wrote ``The Elements'', the most well 
known systematically textbook of the Greek mathematics.} 
knew this in the third century before Christ.
\begin{theorem}[Euclid\footnote{The common usage of the term does not denote Euclid as the inventor of the theorem rather;
the true inventor is merely not as prominent. The theorem has already been distinguished
and proven in Euclidís Elements (Book IX, s. 20). The phraseology is remarkable due to 
the fact that the word infinite is not used. The text reads as followed
$$
O\acute{\iota}~\pi\varrho\tilde{\omega}\tau o \iota~\grave{\alpha}\varrho\iota\vartheta\mu o\grave{\iota}~
\pi\lambda\varepsilon\acute{\iota}o \upsilon\varsigma~\varepsilon\grave{\iota}\sigma\grave{\iota}~
\pi\alpha\nu\tau\grave{o}\varsigma~\tau o \tilde{\upsilon}~
\pi\varrho o \tau\varepsilon\vartheta\acute{\varepsilon}\nu\tau o \varsigma~
\pi\lambda\acute{\eta}\vartheta\ o \upsilon\varsigma~
\pi\varrho\acute{\omega}\tau\omega\nu~
\grave{\alpha}\varrho\iota\vartheta\mu\tilde{\omega}\nu,
$$
the English translation of which is: the prime numbers are more than 
any previously existing amount of prime numbers.
}]\label{thm-pz-euklid} % Ende der Fuﬂnote
% Fussnote vor ], damit kein Leerraum vor dem Fussnotennummer
The sequence of prime numbers does not discontinue.
Therefore, the quantity of prime numbers is infinite.
\end{theorem}
His proof that there is an infinite number of
primes is still considered to be a brilliant mathematical consideration and
conclusion today (proof by contradiction). He assumed that there is only a
finite number of primes and therefore a largest prime number. Based on this
assumption, he drew logical conclusions until he obtained an obvious
contradiction. This meant that something must be wrong. As there were no
mistakes in the chain of conclusions, it could only be the assumption that was
wrong. Therefore, there must be an infinite number of primes!

\hypertarget{euclid}{}
\paragraph{Euclid's proof by contradiction} argues as follows:

@@

Thus there is an {\em infinite} number of primes
\hyperlink{primhfk}{(Cross-reference: overview under \ref{s:primhfk} of the number of
prime numbers in various intervals)}.\par \vskip + 3pt

Here we may mention yet another fact which is initially somewhat surprising. 
Namely, in the prime numbers sequence $p_1, p_2, \cdots,$ gaps between prime numbers can have
an individually determined length $n$. It is undeniable that under the $n$
succession of natural numbers
$$(n+1)!+2,\cdots, (n+1)!+(n+1),
$$
none of them is a prime number since in order, the numbers $2,3,\cdots,(n+1)$  
are comprised respectively as real divisors. 
($n!$ means the product of the first $n$ natural numbers therefore 
$ n!= n*(n-1)*\cdots *2*1$).

@@

% --------------------------------------------------------------------------
\hypertarget{MersenneNumbers01}{}
\subsubsection{Special number types -- Mersenne numbers} \index{Mersenne!Mersenne numbers}

@@

\index{Number!Mersenne}\index{Mersenne!Mersenne!number}
\index{Mersenne!theorem}

Even Mersenne knew that not all Mersenne numbers
are prime (see exponent $p = 11$). 
A prime Mersenne number \index{Mersenne!Mersenne prime number} is called Mersenne prime number.  \\
However, he is to be thanked for the interesting conclusion that a number of the form $2^n-1$ is not a prime number
if $n$ is a composite number:

@@

Due to the fact that the exponents of the Mersenne numbers
\index{Mersenne!number} do not use all
natural numbers, but only the primes, the {\em experimental space} is limited
considerably. The currently known Mersenne prime numbers 
\index{Mersenne!Mersenne prime number} have the exponents 
$$
\begin{array}{c}
2; ~ 3; ~ 5; ~ 7; ~ 13; ~ 17; ~ 19; ~ 31; ~ 61; ~ 89; ~ 107; ~ 127;
~ 521; ~ 607; ~ 1,279; ~ 2,203; ~ 2,281; ~ 3,217; ~ 4,253;\\
 4,423; ~ 9,689; ~ 9,941, ~ 11,213; ~ 19,937; ~ 21,701; ~ 23,207; ~ 44,497; ~
86,243; ~ 110,503; ~ 132,049; \\
 216,091; ~ 756,839; ~ 859,433; ~ 1,257,787; ~ 1,398,269; ~ 2,976,221; ~ 3,021,377; ~
6,972,593; \\13,466,917.
\end{array}
$$
Thus $39$ Mersenne prime numbers are currently known
\index{Prime number!Mersenne}\index{Mersenne!Mersenne prime number}. 
For the first 38 Mersenne prime numbers we know that this list is complete. The exponents until the 39th Mersenne prime numbers are not yet checked completely.

@@

\vskip +15 pt
\paragraph{M-37 -- January 1998}\index{Mersenne!Mersenne prime number!M-37}\mbox{}
\vskip +10 pt

The 37th Mersenne prime,
$$ 2^{3,021,377} - 1. $$
was found in January 1998 and has 909,526
digits in the decimal system, which corresponds to 33 pages in the newspaper!

\vskip +15 pt
\paragraph{M-38 -- June 1999}\index{Mersenne!Mersenne prime number!M-38}\mbox{}
\vskip +10 pt

The 38th Mersenne prime, called M-38,
$$ 2^{6,972,593} - 1. $$
was discovered in June 1999 and has $2,098,960$ digits in the decimal system
(that corresponds to around 77 pages in the newspaper).

Right now (Feb. 2003) all exponents smaller than $ 6,972,593 $ have been
checked: so it is sure, that this is really the 38th Mersenne prime number.
This means there is no other Mersenne prime number between the 37th 
Mersenne prime number and this number.

\vskip +15 pt
\paragraph{M13466917 -- December 2001} \index{Mersenne!Mersenne prime number!M-39}\mbox{}
\vskip +10 pt

This number was discovered as 39th Mersenne prime (and already called M-39, despite it is not proven yet, whether there do not exist further Mersenne prime numbers between M-38 und M13466917), $$2^{13,466,917}-1,$$ at December 6, 2001 ñ- more exactly, the verification of this number, found at November 14, 2001 by the Canadian student Michael Cameron, was successfully finished. This number has about 4 million decimal digits (exactly 4,053,946 digits).
Trying only to print this number 
$$(924947738006701322247758 \cdots 1130073855470256259071)$$
would need around 200 pages in the Financial Times.

@@

% --------------------------------------------------------------------------
\vskip +15 pt
\subsubsection{EFF}\index{EFF}
This search is also spurred on by a competition started by the non-profit
organisation EFF (Electronic Frontier Foundation) using the means of an unknown
donator. The participants are rewarded with a total of 500,000 USD if they find
the longest prime number. In promoting this project, the unknown donator is not
looking for the quickest computer, but rather wants to draw people's attention
to the possibilities offered by {\em cooperative networking} \\
{\href{http://www.eff.org/coopawards/prime-release1.html}{\tt http://www.eff.org/coopawards/prime-release1.html}}

The discoverer of M-38 got for discovering the first prime with more than 1 million decimal digits 50,000 USD
from the EFF. The next price from EFF with 100,000 USD is for a proven prime with more than 10 million decimal digits.
%{\href{http://www.octocad.demon.co.uk/mersenne/prime.htm }{\tt http://www.octocad.demon.co.uk/mersenne/prime.htm}}.

@@

Up to now we have always looked for the prime factors in order to decide whether
a number is prime. However, if even the smallest prime factor is enormous, the
search takes too long. Factorising numbers using systematic computational
division or using the sieve of Eratosthenes (see below) is only feasible using
current computers for numbers with up to around $20$ digits in the decimal
system.

@@

\hypertarget{FermatNumbers01}{}
In the 17th century, Fermat wrote to Mersenne \index{Mersenne Marin} that he presumed that all numbers
of the form $$ F(n) = 2^{2^n} + 1 $$ are prime for all whole numbers $n$ greater
than or equal to $0$ (\hyperlink{FermatNumbers02}{see below}) 
\index{Number!Fermat number}\index{Fermat!Fermat number}

@@

Ideally, we would like, for the number $n$, to immediately obtain the $n$-th
prime number, i.e. for $f(8) = 19\,$ or for  $f(52) = 239$.

@@

    At February 22, 2003 John Cosgrave discovered 
    \begin{itemize} 
     \item the largest composite Fermat number yet and
     \item the largest prime non-Mersenne prime yet with 645,817 digits.
    \end{itemize}

@@

\vskip +15 pt
\hypertarget{generalizedFermatprimes}{}
\item {\em Generalized Fermat numbers}\footnote{%
The base of this power is no more restricted to 2 !}  $F(b,m) = b^{2^m} + 1$:
\index{Fermat!Fermat number!Generalized}  \\
    Generalized Fermat numbers are more numerous than Mersenne numbers at
    equal size and many of them are waiting to be discovered to fill the
    big gaps between the Mersenne primes already found or still undiscovered.
    Progress in number theory made it possible that numbers, where the
    representation is not limited to the base 2, can be tested at almost
    the same speed than a Mersenne number.

@@

$$
 \begin{array}{ll}
f(0) = 41 & \mapsto {\rm ~prime} \\
f(1) = 43 & \mapsto {\rm ~prime} \\
f(2) = 47 & \mapsto {\rm ~prim} \\
f(3) = 53 & \mapsto {\rm ~prime} \\
f(4) = 61 & \mapsto {\rm ~prime} \\
f(5) = 71 & \mapsto {\rm ~prime} \\
f(6) = 83 & \mapsto {\rm ~prime} \\
f(7) = 97 & \mapsto {\rm ~prime} \\
\vdots \\
f(39) = 1,601 & \mapsto {\rm ~prime} \\
f(40) = 11 \cdot 151 &  \mapsto {\rm ~NOT~prime}! \\
f(41) = 41 \cdot 43 &   \mapsto {\rm ~NOT~prime}! \\
\end{array} 
$$

@@

    This means there is no hope to look for further formulas 
    similiar to that of type 11. or 12.

@@

These formulae, which only apply when n tends to infinity, can be replaced by
more precise formulae.
For $x \geq 67$:
$$ ln(x) - 1,5 < x / PI(x) < ln(x) - 0,5 $$
As we know that $PI(x)  =  x / \ln x$ only for very large $x$ ($x$ tending
towards infinity), we can create the following overview:
$$
\begin{array}{ccccc}
x     &  ln(x)  &  x / ln(x) & PI(x)(counted) &       PI(x) / (x/ln(x)) \\
10^3  &  6.908  &   144      &  168        &       1.160 \\
10^6  &  13.816 &   72,386    &  78,498          &       1.085 \\
10^9  & 20.723  &   48,254,942 &  50,847,534       &       1.054
\end{array}
$$

@@

The \textbf{distribution} of primes displays several irregularities for which no
ÄsystemÄ has yet been found: On the one hand, many occur closely together, like
$2$ and $3,$ $ 11$ and $13, $ $ 809$ and $811$, on the other hand large gaps
containing no primes also occur. For example, no primes lie between $113$ and
$127, $ $ 293$ and $307, $ $317$ and $331, $ $ 523$ and $541, $ $ 773$ and $787,
$ $ 839$ and $853$ as well as between $887$ and $907$. \\
For details, see: 
\vspace{-10pt}
\begin{itemize}
  \item[] \href{http://www.utm.edu/research/primes/notes/gaps.html}
               {\tt http://www.utm.edu/research/primes/notes/gaps.html}
\end{itemize}

@@

An easy way of calculating all $PI(x)$ primes less than or equal to $x$ is to
use the sieve of Eratosthenes. In the 3rd century before Christ, he found an
extremely easy, automatic way of finding this out. To begin with, you write down
all numbers from 2 to $x$, circle 2, then cross out all multiples of 2. Next,
you circle the lowest number that hasn't been circled or crossed out (3) and
again cross out all multiples of this number, etc. You only need to continue
until you reach the largest number whose square is less than or equal to $x$.

@@

Details of this can be found under:
\vspace{-10pt}
\begin{itemize}
  \item[] \href{http://www.cerias.purdue.edu/homes/ssw/cun}
               {\tt http://www.cerias.purdue.edu/homes/ssw/cun}
\end{itemize}

@@

\begin{itemize}
  \item For each number $n$ in ${\bf N}$ there are $n$ consecutive natural
numbers that are not primes.
    A proof of this can be found in \cite[p. 79]{Padberg1996}.
  \item The Hungarian mathematician Paul Erd\"os (1913-1996) proved:
    Between each random number not equal to $1$ and its double, there is at
least one prime. (He was not the first to prove this theorem, but proved it in a
simpler manner than those before him).
  \item There is a real number a such that the function $f: {\bf N} \rightarrow
{\mathbb Z}$ where $n \mapsto a^{3^n}$     only delivers primes for all $n$ (see
\cite[p. 82]{Padberg1996}).
    Unfortunately, problems arise when we try to determine $a$ (see below).
\end{itemize}

@@

Twin primes are prime numbers whose difference is 2. Examples include 5 and 7 or
101 and 103. Triplet primes, however, only occur once: 3, 5, 7.  For all other
sets of three consecutive uneven numbers, one of them is always divisible by 3
and thus not a prime.
\begin{itemize}
  \item The number of twin primes is an open question: infinite or limited
number?
  The largest twin primes known today are $1,693,965 \cdot 2^{66,443} \pm 1.$
  \item Does a formula exist for calculating the number of twin primes per
interval?
  \item The above proof of the function  $f: \: N \rightarrow Z$ with $n \mapsto
a^{3^n}$  only guarantees the existence of such a number $a$.  How can we
determine this number $a$ and will it have a value, making the function also of
practical interest?
\item Is there an infinite number of Mersenne prime numbers?\index{Prime number!Mersenne}\index{Mersenne!Mersenne prime number}
\item Is there an infinite number of Fermat prime numbers?

@@

\bibitem[Aaronson2003]{Aaronson2003} \index{Aaronson 2003}
    Scott Aaronson, \\
    {\em The Prime Facts: From Euclid to AKS}, \\
    \href{http://www.cs.berkeley.edu/~aaronson/prime.ps}
         {\texttt{http://www.cs.berkeley.edu/~aaronson/prime.ps}}. \\
    Only after I had completed this article, I did come across the 
    beautiful paper of von Scott Aaronson, which also offers a simple, 
    didactically well done but nervertheless not to low aimed intoduction
    into this topic.

@@

\begin{enumerate}
\item GIMPS (Great Internet Mersenne-Prime Search) 
      \index{Mersenne!Mersenne prime number} \\
      www.mersenne.org is the home page of the GIMPS project, \index{GIMPS} \\
      \href{http://www.mersenne.org/prime.htm}{\tt http://www.mersenne.org/prime.htm }

@@

\item At the university of Tennessee you will find extensive research
      results about prime numbers. \\
      \href{http://www.utm.edu/}{\tt http://www.utm.edu/ }

@@

I would like to take this opportunity to thank Mr. Henrik Koy and Mr. Roger
Oyono for their very constructively proof-reading of this article.

>>>>

Prime numbers are particularly important for 3 reasons:
\begin{itemize}
  \item In number theory, they are considered to be the basic components of
natural numbers, upon which numerous brilliant mathematical ideas are based.
  \item They are of extreme practical importance in modern
\index{Cryptography!modern} cryptography (public key \index{Cryptography!public key} cryptography). The most common public key procedure, invented at the end of
the 1970's, is \index{RSA} RSA encryption. Only using (large) prime numbers for
particular parameters can you guarantee that an algorithm is secure, both for
the RSA procedure and for even more modern procedures (digital
\index{Signatures!digital} signature, elliptic curves).
  \item The search for the largest known prime numbers does not have any
practical usage known to date, but requires the best computers, is an excellent
benchmark (possibility for determining the performance\index{Performance} of computers) and leads
to new calculation methods on many computers \\ (see also:
\href{http://www.mersenne.org/prime.htm}{\tt
http://www.mersenne.org/prime.htm}).
\end{itemize}
Many people have been fascinated by prime numbers over the past two millennia.
Ambition to make new discoveries about prime numbers has often resulted in
brilliant ideas and conclusions. The following section provides an easily
comprehensible introduction to the basics of prime numbers. We will also explain
what is known about the distribution (density, number of prime numbers in
particular intervals) of prime numbers and how prime number tests work.

@@

The sequence of prime numbers starts with $$ 2,~ 3,~ 5,~ 7, ~ 11, ~ 13, ~
17, ~ 19, ~ 23, ~ 29, ~ 31, ~ 37, ~ 41, ~ 43, ~ 47, ~ 53, ~ 59, ~ 61,
~ 67, ~ 71, ~ 73, ~ 79, ~ 83, ~ 89, ~ 97, \cdots . $$
The first 100 numbers include precisely 25 prime numbers. After this,
the percentage of primes constantly decreases. Prime numbers can be
factorised in a uniquely {\em trivial} way: 
$$5 = 1 \cdot 5,\quad  17 = 1 \cdot 17, \quad 1,013 = 1 \cdot 1,013,  \quad
1,296,409 = 1 \cdot 1,296,409.$$
All numbers that have $2$ or more factors are called \index{Number!composite}
{\em composite}. These include $$ 4 = 2 \cdot 2, \quad 6 = 2\cdot 3 $$ as well
as numbers that {\em look like primes}, but are in fact composite:
$$ 91 = 7 \cdot 13, \quad 161=7 \cdot 23, \quad 767 =13 \cdot 59. $$

@@

Although there are only $92$ natural chemical elements, the number of prime
numbers is unlimited. Even the Greek, \index{Euclid} Euclid%
\footnote{Euclid,
a Greek mathematician of 4th and 3rd century B.C. He worked at the
Egyptian academy of Alexandria and wrote ``The Elements'', the most well 
known systematically textbook of the Greek mathematics.} 
knew this in the third century before Christ.
\begin{theorem}[Euclid\footnote{The common usage of the term does not denote Euclid as the inventor of the theorem rather;
the true inventor is merely not as prominent. The theorem has already been distinguished
and proven in Euclidís Elements (Book IX, s. 20). The phraseology is remarkable due to 
the fact that the word infinite is not used. The text reads as followed
$$
O\acute{\iota}~\pi\varrho\tilde{\omega}\tau o \iota~\grave{\alpha}\varrho\iota\vartheta\mu o\grave{\iota}~
\pi\lambda\varepsilon\acute{\iota}o \upsilon\varsigma~\varepsilon\grave{\iota}\sigma\grave{\iota}~
\pi\alpha\nu\tau\grave{o}\varsigma~\tau o \tilde{\upsilon}~
\pi\varrho o \tau\varepsilon\vartheta\acute{\varepsilon}\nu\tau o \varsigma~
\pi\lambda\acute{\eta}\vartheta\ o \upsilon\varsigma~
\pi\varrho\acute{\omega}\tau\omega\nu~
\grave{\alpha}\varrho\iota\vartheta\mu\tilde{\omega}\nu,
$$
the English translation of which is: the prime numbers are more than 
any previously existing amount of prime numbers.
}]\label{thm-pz-euklid} % Ende der Fuﬂnote
% Fussnote VOR "]" in [Euclid], damit kein Leerraum vor der Fussnotennummer.
% Vorher stand da: ...(Euclid). BLANK Fussnote und das Blank stoerte.
% Nun steht die Fussnote direkt hinter "Euclid" und vor der ")".
% Eigentlich h‰tte ich sie gerne direkt hinter "(Euclid)", noch vor dem 
% automatisch gesetzten Punkt. 
The sequence of prime numbers does not discontinue.
Therefore, the quantity of prime numbers is infinite.
\end{theorem}
His proof that there is an infinite number of
primes is still considered to be a brilliant mathematical consideration and
conclusion today (proof by contradiction). He assumed that there is only a
finite number of primes and therefore a largest prime number. Based on this
assumption, he drew logical conclusions until he obtained an obvious
contradiction. This meant that something must be wrong. As there were no
mistakes in the chain of conclusions, it could only be the assumption that was
wrong. Therefore, there must be an infinite number of primes!

\hypertarget{euclid}{}
\paragraph{Euclid's proof by contradiction} goes as follows:

@@

Thus there is an {\em infinite} number of primes
\hyperlink{primhfk}{(Cross-reference: overview under \ref{s:primhfk} of 
the number of prime numbers in various intervals)}.\par \vskip + 10pt

Here we should perhaps mention yet another fact which is initially somewhat surprising. 
Namely, in the prime numbers sequence $p_1, p_2, \cdots,$ gaps between prime numbers can have
an individually determined length $n$. It is undeniable that under the $n$
succession of natural numbers
$$(n+1)!+2,\cdots, (n+1)!+(n+1),
$$
none of them is a prime number since in order, the numbers $2,3,\cdots,(n+1)$  
are comprised respectively as real divisors. 
($n!$ means the product of the first $n$ natural numbers therefore 
$ n!= n*(n-1)*\cdots *2*1$).

@@

% --------------------------------------------------------------------------
\hypertarget{MersenneNumbers01}{}
\subsubsection{Special number types -- Mersenne numbers} 
\index{Mersenne!number}

@@

\index{Number!Mersenne}\index{Mersenne!number}
\index{Mersenne!theorem}

Even Mersenne knew that not all Mersenne numbers
are prime (see exponent $p = 11$). 
A prime Mersenne number \index{Mersenne!prime number} is called
Mersenne prime number.  \\
However, he is to be thanked for the interesting conclusion that a number of the form $2^n-1$ is not a prime number
if $n$ is a composite number:

@@

Due to the fact that the exponents of the Mersenne numbers
\index{Mersenne!number} do not use all
natural numbers, but only the primes, the {\em experimental space} is limited
considerably. The currently known Mersenne prime numbers 
\index{Mersenne!prime number} have the exponents 
$$
\begin{array}{c}
2; ~ 3; ~ 5; ~ 7; ~ 13; ~ 17; ~ 19; ~ 31; ~ 61; ~ 89; ~ 107; ~ 127;
~ 521; ~ 607; ~ 1,279; ~ 2,203; ~ 2,281; ~ 3,217; ~ 4,253;\\
 4,423; ~ 9,689; ~ 9,941, ~ 11,213; ~ 19,937; ~ 21,701; ~ 23,207; ~ 44,497; ~
86,243; ~ 110,503; ~ 132,049; \\
 216,091; ~ 756,839; ~ 859,433; ~ 1,257,787; ~ 1,398,269; ~ 2,976,221; ~ 3,021,377; ~
6,972,593; \\13,466,917.
\end{array}
$$
Thus $39$ Mersenne prime numbers are currently known
\index{Prime number!Mersenne}\index{Mersenne!prime number}. 
For the first 38 Mersenne prime numbers we know that this list is complete.
The exponents until the 39th Mersenne prime number have not yet been checked
completely.

@@

\vskip +15 pt
\paragraph{M-37 -- January 1998}\index{Mersenne!prime number!M-37}\mbox{}
\vskip +10 pt

The 37th Mersenne prime, $$ 2^{3,021,377} - 1 $$
was found in January 1998 and has 909,526
digits in the decimal system, which corresponds to 33 pages in the newspaper!

\vskip +15 pt
\paragraph{M-38 -- June 1999}\index{Mersenne!prime number!M-38}\mbox{}
\vskip +10 pt

The 38th Mersenne prime, called M-38, $$ 2^{6,972,593} - 1 $$
was discovered in June 1999 and has $2,098,960$ digits in the decimal system
(that corresponds to around 77 pages in the newspaper).

Right now (Feb. 2003) all exponents smaller than $ 6,972,593 $ have been
checked: so we can be certain, that this is really the 38th Mersenne prime
number.
This means there is no other Mersenne prime number between the 37th 
Mersenne prime number and this one.

\vskip +15 pt
\paragraph{M13466917 -- December 2001} \index{Mersenne!prime number!M-39}\mbox{}
\vskip +10 pt

This number was discovered as 39th Mersenne prime (and already called M-39,
despite it has not been proven yet, whether no further Mersenne prime
numbers between M-38 und M13466917 do indeed exist), $$2^{13,466,917}-1,$$
at December 6, 2001 -- more exactly, the verification of this number,
found at November 14, 2001 by the Canadian student Michael Cameron, was
successfully completed. This number has about 4 million decimal digits
(exactly 4,053,946 digits).
Trying only to print this number 
$$(924947738006701322247758 \cdots 1130073855470256259071)$$
would need around 200 pages in the Financial Times.

@@

% --------------------------------------------------------------------------
\vskip +15 pt
\subsubsection{EFF}\index{EFF}
This search is also spurred on by a competition started by the non-profit
organisation EFF (Electronic Frontier Foundation) using the means of an unknown
donator. The participants are rewarded with a total of 500,000 USD if they find
the longest prime number. In promoting this project, the unknown donator is not
looking for the quickest computer, but rather wants to draw people's attention
to the opportunities offered by {\em cooperative networking} \\
{\href{http://www.eff.org/coopawards/prime-release1.html}{\tt http://www.eff.org/coopawards/prime-release1.html}}

The discoverer of M-38 received 50,000 USD from the EFF for discovering 
the first prime with more than 1 million decimal digits. 
The next prize from EFF for 100,000 USD is for a proven prime with more
than 10 million decimal digits.
%{\href{http://www.octocad.demon.co.uk/mersenne/prime.htm }{\tt http://www.octocad.demon.co.uk/mersenne/prime.htm}}.

@@

Up to now we have always looked for the prime factors in order to decide whether
a number is prime. However, even if the smallest prime factor is enormous, the
search takes too long. Factorising numbers using systematic computational
division or using the sieve of Eratosthenes (see below) is only feasible using
current computers for numbers with up to around $20$ digits in the decimal
system.

@@

\hypertarget{FermatNumbers01}{}
In the 17th century, Fermat wrote to Mersenne \index{Mersenne Marin} that he presumed that all numbers
of the form $$ F(n) = 2^{2^n} + 1 $$ are prime for all whole numbers $n$ greater
than or equal to $0$ (\hyperlink{FermatNumbers02}{see below}) 
\index{Number!Fermat}\index{Fermat!number}

@@

Ideally, we would like, for the number $n$, to immediately be able to 
obtain the $n$-th prime number, i.e. for $f(8) = 19\,$ or for  $f(52) = 239$.

@@

    On February 22, 2003 John Cosgrave discovered 
    \begin{itemize} 
     \item the largest composite Fermat number to date and
     \item the largest prime non-Mersenne number so far with 645,817 digits.
    \end{itemize}

@@

\vskip +15 pt
\hypertarget{generalizedFermatprimes}{}
\item {\em Generalized Fermat numbers}\footnote{%
The base of this power is no longer restricted to 2 !}  $F(b,m) = b^{2^m} + 1$:
\index{Fermat!Fermat number!Generalized}  \\
    Generalized Fermat numbers are more numerous than Mersenne numbers at
    equal size and many of them are waiting to be discovered to fill the
    big gaps between the Mersenne primes already found or still undiscovered.
    Progress in number theory made it possible that numbers, where the
    representation is not limited to the base 2, can be tested at almost
    the same speed than a Mersenne number.

@@

$$
 \begin{array}{ll}
f(0) = 41 & \mapsto {\rm ~prime} \\
f(1) = 43 & \mapsto {\rm ~prime} \\
f(2) = 47 & \mapsto {\rm ~prime} \\
f(3) = 53 & \mapsto {\rm ~prime} \\
f(4) = 61 & \mapsto {\rm ~prime} \\
f(5) = 71 & \mapsto {\rm ~prime} \\
f(6) = 83 & \mapsto {\rm ~prime} \\
f(7) = 97 & \mapsto {\rm ~prime} \\
\vdots \\
f(39) = 1,601 & \mapsto {\rm ~prime} \\
f(40) = 11 \cdot 151 &  \mapsto {\rm ~NOT~prime}! \\
f(41) = 41 \cdot 43 &   \mapsto {\rm ~NOT~prime}! \\
\end{array} 
$$

@@

    This means there is no hope in looking for further formulae 
    similar to that of type 11. or 12.

@@

These formulae, which only apply when n tends to infinity, can be replaced by
more precise formulae.
For $x \geq 67$:
$$ ln(x) - 1,5 < x / PI(x) < ln(x) - 0,5 $$
Given that we know $PI(x)  =  x / \ln x$ only for very large $x$ ($x$ tending
towards infinity), we can create the following overview:
$$
\begin{array}{ccccc}
x     &  ln(x)  &  x / ln(x) & PI(x)(counted) &       PI(x) / (x/ln(x)) \\
10^3  &  6.908  &   144      &  168        &       1.160 \\
10^6  &  13.816 &   72,386    &  78,498          &       1.085 \\
10^9  & 20.723  &   48,254,942 &  50,847,534       &       1.054
\end{array}
$$

@@

The \textbf{distribution} of primes displays several irregularities for which no
ÄsystemÄ has yet been found: On the one hand, many occur closely together, like
$2$ and $3,$ $ 11$ and $13, $ $ 809$ and $811$, on the other hand large gaps
containing no primes also occur. For example, no primes lie between $113$ and
$127, $ $ 293$ and $307, $ $317$ and $331, $ $ 523$ and $541, $ $ 773$ and $787,
$ $ 839$ and $853$ as well as between $887$ and $907$. \\
For details, please see: 
\vspace{-10pt}
\begin{itemize}
  \item[] \href{http://www.utm.edu/research/primes/notes/gaps.html}
               {\tt http://www.utm.edu/research/primes/notes/gaps.html}
\end{itemize}

@@

An easy way of calculating all $PI(x)$ primes less than or equal to $x$ is to
use the sieve of Eratosthenes. In the 3rd century B.C., he found an
extremely easy, automatic way of finding this out. To begin with, you write down
all numbers from 2 to $x$, circle 2, then cross out all multiples of 2. Next,
you circle the lowest number that hasn't been circled or crossed out (3) and
again cross out all multiples of this number, etc. You only need to continue
until you reach the largest number whose square is less than or equal to $x$.

@@

Details of this can be found at:
\vspace{-10pt}
\begin{itemize}
  \item[] \href{http://www.cerias.purdue.edu/homes/ssw/cun}
               {\tt http://www.cerias.purdue.edu/homes/ssw/cun}
\end{itemize}

@@

\begin{itemize}
  \item For each number $n$ in ${\bf N}$ there are $n$ consecutive natural
      numbers that are not primes.
      A proof of this can be found in \cite[p. 79]{Padberg1996}.
  \item The Hungarian mathematician Paul Erd\"os (1913-1996) proved:
      Between each random number not equal to $1$ and its double, there is
      at least one prime. He was not the first to prove this theorem, but
      proved it in a much simpler manner than those before him.
  \item There is a real number a such that the function $f: {\bf N}
      \rightarrow {\mathbb Z}$ where $n \mapsto a^{3^n}$ 
      only delivers primes for all $n$ (see \cite[p. 82]{Padberg1996}).
      Unfortunately, problems arise when we try to determine $a$ (see below).
\end{itemize}

@@

Twin primes are prime numbers whose difference is 2. Examples include 5 and 7 or
101 and 103. Triplet primes, however, only occur once: 3, 5, 7.  For all other
sets of three consecutive uneven numbers, one of them is always divisible by 3
and thus not a prime.
\begin{itemize}
  \item The number of twin primes is an open question: infinite or limited
number?
  The largest twin primes known today are $1,693,965 \cdot 2^{66,443} \pm 1.$
  \item Does a formula exist for calculating the number of twin primes per
interval?
  \item The above proof of the function  $f: \: N \rightarrow Z$ with $n \mapsto
a^{3^n}$  only guarantees the existence of such a number $a$.  How can we
determine this number $a$ and will it have a value, making the function also of
some practical interest?
\item Is there an infinite number of Mersenne prime numbers?\index{Prime number!Mersenne}\index{Mersenne!prime number}
\item Is there an infinite number of Fermat prime numbers?

@@

\bibitem[Aaronson2003]{Aaronson2003} \index{Aaronson 2003}
    Scott Aaronson, \\
    {\em The Prime Facts: From Euclid to AKS}, \\
    \href{http://www.cs.berkeley.edu/~aaronson/prime.ps}
         {\texttt{http://www.cs.berkeley.edu/~aaronson/prime.ps}}. \\
    Only after I had completed this article, I did come across the 
    extremely well-written paper by Scott Aaronson, which also offers
    a didactically well done intoduction to this topic. It is 
    humorous and easy to read but at the same time precise and erudite.

@@

\begin{enumerate}
\item GIMPS (Great Internet Mersenne-Prime Search) 
      \index{Mersenne!prime number}  \index{GIMPS} \\
      www.mersenne.org is the home page of the GIMPS project, \\
      \href{http://www.mersenne.org/prime.htm}{\tt http://www.mersenne.org/prime.htm }

@@

\item At the University of Tennessee you will find extensive research
      results about prime numbers. \\
      \href{http://www.utm.edu/}{\tt http://www.utm.edu/ }

@@

I would like to take this opportunity to thank Mr. Henrik Koy and Mr. Roger
Oyono for their very constructive proof-reading of this article.
